---
title: "Waikato Transport Analysis"
output: html_notebook
---

- split map jobs function so that the households and pop per zone are calculated, then this intermediate output can be used for all the other census variables rather than loading in the address data everytime
- or can i just use the jobs_h3 output anyway?
#load in previous analysis
```{r}
#ttm data
ttm_bicycle_results_list <- readRDS("ttm_bicycle_results_list.rds")

#access to jobs and education
walk_results <- readRDS("walk_results.rds")
cycling_results <- readRDS() 
transit_results <- readRDS()
car_results <- readRDS()
```
# load packages and functions -----------------------------------------
```{r}
#increases memory available to java
options(java.parameters = "-Xmx4G")

# install dev version with latest features
#devtools::install_github("ipeaGIT/r5r", subdir = "r-package")

library(r5r)
library(sf)
library(data.table)
library(interp)
library(tmap)
library(dplyr)
library(accessibility)
#load additional packages
library (osmdata)
library(osmextract)
library(h3jsr)
library(purrr)
library(ggplot2)
packageVersion ("osmdata")
source("C:\\Users\\ellio\\OneDrive\\Documents\\GIS Data\\R5R\\my_functions.r")
#https://ipeagit.github.io/r5r/articles/accessibility.html
tmap_mode("view")
```



# define area of analysis ---------------------

#Create bounding box for townships
```{r}
#http://bboxfinder.com/
# Create a matrix with the bounding box coordinates and ensure it is closed
waikato_bbox <- create_bounding_box(174.987992,-37.513760, 120)
#tm_shape(waikato_bbox) + tm_polygons(alpha = 0.5)

#create all the bbox's
raglan_bbox <- tidygeocoder::geo(address = "Raglan, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))() #\(df) defines a function and () at the end call it 

pokeno_bbox <- tidygeocoder::geo(address = "Pokeno, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

port_waikato_bbox <- tidygeocoder::geo(address = "Port Waikato, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

horotiu_bbox <- tidygeocoder::geo(address = "Horotiu, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

meremere_bbox <- tidygeocoder::geo(address = "Meremere, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

tuakau_bbox <- tidygeocoder::geo(address = "Tuakau, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

tekauwhata_bbox <- tidygeocoder::geo(address = "Te Kauwhata, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

huntly_bbox <- tidygeocoder::geo(address = "Huntly, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

taupiri_bbox <- tidygeocoder::geo(address = "Taupiri, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

ngaruawahia_bbox <- tidygeocoder::geo(address = "Ngaruawahia, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

maramarua_bbox <- tidygeocoder::geo(address = "Maramarua, Waikato, New Zealand") |> 
  (\(df) create_bounding_box(df$long, df$lat, size_km = 10))()

bbox_list <- list(raglan_bbox = raglan_bbox, pokeno_bbox = pokeno_bbox, port_waikato_bbox = port_waikato_bbox, horotiu_bbox = horotiu_bbox, meremere_bbox = meremere_bbox, tuakau_bbox = tuakau_bbox, tekauwhata_bbox = tekauwhata_bbox, huntly_bbox = huntly_bbox, taupiri_bbox = taupiri_bbox, ngaruawahia_bbox = ngaruawahia_bbox, maramarua_bbox = maramarua_bbox)

combined_bboxs <- bind_rows(bbox_list, .id = "area_name")

#visually inspect
#tm_shape(combined_bboxs) + tm_polygons(alpha = 0.5)

```

#create additional bounding box for analysis of wider area
```{r}
#reuse existing bbox's
#Find centroids
bbox_centroids <- map(bbox_list, ~ st_centroid(st_union(.x)) |> st_coordinates()) 

#create bounding boxes from previously definied custom function
bbox_list_30km <- map(bbox_centroids, ~ create_bounding_box(.x[1], .x[2], size_km = 20))

#bind together into a single df
combined_bboxs_30km <- bind_rows(bbox_list_30km, .id = "area_name")

#visually inspect
#tmap_study_areas_OD <- tm_shape(combined_bboxs_30km) + 
  #tm_polygons(alpha = 0.5, border.col = "black") + tm_shape(combined_bboxs) + tm_polygons(alpha = 0.5, border.col = "darkgreen") + tm_layout(title = "Waikato Study Area (black = destinations, green = origins")

#tmap_save(tmap_study_areas_OD, "Waikato Study Areas.html")

```


#global parameters -----------------------------------------

Routing inputs and modes
```{r}
# routing inputs
mode <- c("TRANSIT")
max_walk_time <- 45      # in minutes
max_trip_duration <- 60 # in minutes
time_window <- 60       # in minutes
departure_datetime <- as.POSIXct("25-03-2025 08:30:00",
                                 format = "%d-%m-%Y %H:%M:%S")
walk_speed <- 5.1
time_intervals <- seq(0, 100, 10)
#res 8 for cars res 10 for everything else
res <- 10
amenities <- c("school", "hospital", "university", "bus_station", "charging_station", "dentist", "doctors", "police", "post_office", "toilets")
```

# Load in opportunities data

    
```{r}
#amenity data
opportunities <- download_opq_opportunities(waikato_bbox |> st_bbox(), amenities = amenities, res = res, place_name = "Waikato")
```

#ensure land use data ids align to ttm so that there are no incorrect NA's passed

Without this the travel times may come back as NA's rather than the actual times.
```{r}
#add additional zone id's
opportunities$combined_summary_all <- opportunities$summary |>
  rows_insert(waikato_routing_cells, conflict = "ignore") |> #ignore means nonconflicted rows are added
  mutate(across(
    where(is.numeric), #only apply to numeric columns
    ~ tidyr::replace_na(., 0)
  ))
#checks
opportunities$combined_summary_all |> View()
opportunities$combined_summary_all |> ncol()
opportunities$summary |> ncol()

#check data is identical
opportunities$combined_summary_all %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) == opportunities$combined_summary %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
```

#Load Census data
```{r}
#load saved data or rerun?
read_save_location <- "C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato"
  
#custom function to read in and subset data
census_subset <- load_census_data(load_or_recreate_census = "load", study_area = waikato_bbox, read_save_location = read_save_location)
```

#load resident data
```{r}
#load data
#census_subset_resident <- readRDS("census_waikato_resident.rds")

#filter data for resident
census_subset_resident <- census_subset %>%
  filter(CEN23_TBT_IND_003..Variable.codes == "egTotal: Ethnicity (total responses) - total census usually resident population count" &
           CEN23_YEAR_001..Census.year == 2023 &
           OBS_VALUE < 4993922) %>%
  na.omit()

# save the object
saveRDS(census_subset_resident, "census_waikato_resident.rds")
```

#visual inspection of census population
```{r}
tm_shape(census_subset_resident |> st_as_sf()) + tm_polygons(col = "OBS_VALUE", alpha = 0.5)
```

#extract jobs data
```{r}
census_jobs <- census_subset |> filter(CEN23_TBT_IND_003..Variable.codes == "inwTotal: Industry, by workplace address - total employed census usually resident population count aged 15 years and over" &
           CEN23_YEAR_001..Census.year == 2023)
```

#Map the job data to the hex grid
```{r}

map_jobs_to_h3 <- function(load_or_recreate_census = "recreate", census_jobs, study_area, res, read_save_location) {
  # Construct path to saved file (in case only a directory is passed)
  saved_file_path <- file.path(read_save_location, "census_jobs_mapped.rds")
  
  census_jobs <- census_jobs |> st_as_sf()
  
  # Read all NZ address data and crop to study area
  addresses <- st_read("C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Pokeno/lds-nz-addresses-SHP/nz-addresses.shp") %>%
    st_crop(study_area)
  
  # Join addresses to census job zones
  addresses_within <- addresses |>
    st_join(census_jobs, join = st_within)
  
  # Process address counts and jobs allocation
  addresses_within_2 <- addresses_within |> 
    group_by(SA12023_V1) |> 
    mutate(dwellings_in_SA1 = n()) |> 
    ungroup() |> 
    mutate(
      jobs_per_dwelling_SA1 = OBS_VALUE / dwellings_in_SA1,
      h3_address = point_to_cell(geometry, res = res, simple = TRUE)
    ) |> 
    st_drop_geometry() |>  # Drop original point geometries before aggregation
    group_by(h3_address) |> 
    summarise(
      dwellings_h3 = n(),
      jobs = mean(jobs_per_dwelling_SA1, na.rm = TRUE) * mean(dwellings_h3),
      .groups = "drop"
    ) |> 
    cell_to_polygon(simple = FALSE) |> 
    st_as_sf()
  
  # Create tmap
  tmap_jobs <- tm_shape(addresses_within_2) + 
    tm_polygons(
      col = "jobs",
      title = "Jobs",
      alpha = 0.5,
      lwd = 0,
      palette = "YlGn",
      style = "cont",
      popup.vars = c("jobs", "dwellings_h3")
    ) +
    tm_layout(title = "Jobs")

  # Save outputs
  saveRDS(addresses_within_2, saved_file_path)
  
  message(paste("Job data saved to", saved_file_path))
  
  return(list(jobs_data = addresses_within_2, map = tmap_jobs))
}


#custom function to map jobs to h3 grid
jobs_h3 <- map_jobs_to_h3(load_or_recreate_census = "load", census_jobs = census_jobs, study_area = waikato_bbox, res = res, read_save_location = read_save_location)

jobs_h3$jobs_data$h3_address[1] |> get_res()

#visually inspect original census job data
tm_shape(census_jobs |> st_as_sf()) + tm_polygons(col = "OBS_VALUE", alpha = 0.5, palette = "YlGn") + jobs_h3$map

#save map
tmap_save(tm_shape(census_jobs |> st_as_sf()) + tm_polygons(col = "OBS_VALUE", alpha = 0.5, palette = "YlGn") + jobs_h3$map, "Waikato Jobs (Census and h3).html")

```

#map resident population to hex grid
```{r}
map_population_to_h3 <- function(load_or_recreate_census = "recreate", census_pop, study_area, res, read_save_location) {
  # Construct path to saved file (in case only a directory is passed)
  saved_file_path <- file.path(read_save_location, "census_pop_mapped.rds")
  
  census_pop <- census_pop |> st_as_sf()
  
  #read in all NZ address data
  addresses <- st_read("C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Pokeno/lds-nz-addresses-SHP/nz-addresses.shp") %>%
  st_crop(study_area)
  
  #updated code
  addresses_within <-  addresses |> 
  st_join(census_pop, join = st_within)

  addresses_within_2 <- addresses_within |> 
    group_by(SA12023_V1) |> 
    mutate(dwellings_in_SA1 = n()) |> 
    ungroup() |>
    mutate(population_per_dwelling_SA1 = OBS_VALUE / dwellings_in_SA1,
           h3_address = point_to_cell(geometry, res = res, simple = TRUE)) |>
    st_drop_geometry() |> #otherwise combines all point data
    group_by(h3_address) |>
    summarise(dwellings_h3 = n(), 
              population = mean(population_per_dwelling_SA1) * mean(dwellings_h3)) |>
    cell_to_polygon(simple = FALSE)
  
  tmap_pop <- tm_shape(addresses_within_2) + tm_polygons(col = "population", title = "Population", alpha = 0.5, lwd = 0, palette = "YlGn", style = "cont", popup.vars = c("population", "dwellings_h3")) + tm_layout(title = "Population")
  
  saveRDS(addresses_within_2, saved_file_path)
  
  message(paste("pop data saved to ", saved_file_path))
  
  return(list(pop_data = addresses_within_2, map = tmap_pop))
}

#custom function to map pop to h3 grid
pop_h3 <- map_population_to_h3(load_or_recreate_census = "load", census_pop = census_subset_resident, study_area = waikato_bbox, res = res, read_save_location = read_save_location)

tmap_save(pop_h3$map, "Waikato Population H3.html")

#some strange data issues going on here where there seems to be mismatches with the number of dwellings and the number of people

pop_data_head <- pop_h3$pop_data |> arrange(desc(population)) |> slice_head(n = 10) |> st_as_sf()

#check resolution
pop_h3$pop_data$h3_address[1] |> get_res()

# Ensure your input is sf
census_subset_resident <- census_subset_resident %>% st_as_sf()

tmap_census_resident_population <- tm_shape(census_subset_resident) + tm_polygons(alpha = 0.7, col = "OBS_VALUE", palette = "YlGn", title = "Population") + tm_layout(title = "Waikato Census and H3 Population") +
  pop_h3$map

# Find which census features intersect pop_data_head
matches <- st_intersects(census_subset_resident, pop_data_head)

# Keep only those that intersect
high_pop_census <- census_subset_resident[lengths(matches) > 0, ]

matches_address <- st_intersects(pop_h3$addresses, high_pop_census)

high_pop_addresses <-  pop_h3$addresses[lengths(matches_address) > 0,]

matches_h3 <- st_intersects(pop_h3$pop_data, high_pop_census)

high_poph3_absent <- pop_h3$pop_data[lengths(matches_h3) > 0,]
 

tm_shape(pop_data_head) + tm_polygons(alpha = 0.7, col = "population", palette = "YlGn") + tm_shape(high_pop_addresses) + tm_dots() + tm_shape(high_pop_census) + tm_polygons(alpha = 0.7, col = "OBS_VALUE", palette = "YlGn") + tm_shape(high_poph3_absent) + tm_polygons(alpha = 0.7, col = "population", palette = "YlGn", border.col = "blue")

## appears that there are issues either with LINZ not picking up all the addresses, or census allocating too many people to those zones. There doesnt appear to be an issue allocation population.

#visually inspect original census job data
tm_shape(census_jobs |> st_as_sf()) + tm_polygons(col = "OBS_VALUE", alpha = 0.5, palette = "YlGn") + jobs_h3$map

```


#########################################
#extract income data
##########################################
```{r}
#filter data for income
census_subset_median_income <- census_subset %>%
  filter(CEN23_TBT_IND_003..Variable.codes == "ibmed: Median ($) - total personal income" &
           CEN23_YEAR_001..Census.year == 2023) %>%
  na.omit()
```

#map income data
Median income is an average across the zone. No additional work to reallocate based on number of dwellings or population is required
```{r}
#join with jobs data
jobs_population_income <- jobs_h3$jobs_data |> left_join(census_subset_median_income |> 
                                                          select(CEN23_TBT_GEO_002..Area, OBS_VALUE) |> 
                                                          rename(median_income = OBS_VALUE), 
                                                        by = "CEN23_TBT_GEO_002..Area") |>
  left_join(pop_h3$pop_data |> st_drop_geometry() |> select(h3_address, population), by = "h3_address")

#visually inspect median income data
tm_shape(jobs_population_income) + tm_polygons(col = "median_income", alpha = 0.5, lwd = 0, palette = "YlGn")
#doesnt look correct
tm_shape(census_subset_median_income |> st_as_sf()) + tm_polygons(col = "OBS_VALUE", alpha = 0.5, palette = "YlGn")
#there is definitely data for central hamilton
```

# Create routing network for origins. Waikato area h3 id's for routing as origins
```{r}

census_boundaries <- st_read("C:/Users/ellio/OneDrive/Documents/Census 2023/statsnz-statistical-area-1-2023-generalised-SHP") |> st_transform(st_crs(waikato_bbox)) |> st_crop(waikato_bbox) 

#find h3 cells on land
land_cells <- polygon_to_cells(census_boundaries |> filter(LANDWATER_ == "Mainland"), res = res, simple = TRUE) |> 
  list_c() #flatten list
tmap_mode("plot")

#visually inspect 
#land_cells |> cell_to_point() |> tm_shape() + tm_dots()
#census_boundaries|> filter(LANDWATER_ == "Mainland") |> tm_shape() + tm_polygons()
  
#create routing cells for all waikato bbox
all_waikato_routing_cells <- routing_cells(waikato_bbox, res = res)

sub_waikato_routing_cells <- all_waikato_routing_cells[all_waikato_routing_cells$id %in% land_cells,]


#visual inspection
tmap_mode("plot")
#sub_waikato_routing_cells |> tm_shape() + tm_dots()

#assign as waikato_routing_cells
waikato_routing_cells <- sub_waikato_routing_cells

rm(all_waikato_routing_cells, sub_waikato_routing_cells, census_boundaries)
```
################################################################
#walking time matrix and access to jobs and education
#################################################################
Create TTM, undertake access analysis, summarise results and then delete intermediate steps to save working memory.




#Setup core
```{r}

destfile <- "C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato/router_do_min"

# Example: Load R5 network
r5r_core_waikato <- setup_r5(data_path = destfile, overwrite = FALSE)

```


#walking ttm

This creates a many to many matrix of walk times to all other areas. This means the ouputs should also be usable for access to the town centre as well as access to schools, for example.
 
```{r}
#read in saved data
ttm_walk_results_list <- readRDS("ttm_walk_results_list.rds")

#customer function to run ttm for all study areas
run_ttm_for_area <- function(study_area, wider_study_area, area_id, mode) {
  
  #crop destinations 
  cropped_destinations <- waikato_routing_cells[
  st_intersects(waikato_routing_cells, study_area, sparse = FALSE),]
  
  #crop origins
  cropped_origins <- waikato_routing_cells[
  st_intersects(waikato_routing_cells, wider_study_area, sparse = FALSE),]
  
  #catch an issues with origin data
  if (nrow(cropped_destinations) == 0) {
    warning(paste("No origins found for area", area_id))
    return(NULL)
  }
  
  #run travel time matrix
  ttm <- travel_time_matrix(
    r5r_core_waikato,
    origins = cropped_origins,
    destinations = cropped_destinations,
    mode = mode, #change mode independently of global settings
    departure_datetime = departure_datetime,
    max_walk_time = max_walk_time,
    max_trip_duration = max_trip_duration,
    walk_speed = walk_speed
  )
 
  return(ttm)
}

#loop of list of bboxs
ttm_walk_results_list <- purrr::pmap(list(st_geometry(combined_bboxs),
                                          st_geometry(combined_bboxs_30km), 
                                          combined_bboxs$area_name,
                                          c(rep("WALK", 11))),
  .f = run_ttm_for_area) |> set_names(combined_bboxs$area_name)

saveRDS(ttm_walk_results_list, "ttm_walk_results_list.rds")
```

##########################################################
#map and save interactive walk ttm map

Take the centroid of the study areas, which is typically the town centre and crop so that we have walking access to that location. 

This can also be re-used for a number of other locations, if we chant the centroid list to something else.
```{r}
#find centroid h3s
bbox_h3_all <- bbox_list |> bind_rows() |> st_centroid() |> point_to_cell(res = res)

bbox_h3_all <- c(bbox_h3_all,"8abb54726237fff") #additional point for raglan which seems to get cut off

bbox_h3_all <- bbox_h3_all[1:11]

#filter the ttm's based on the locations included in the above
tmap_ttm_walk <- ttm_walk_results_list |> 
  bind_rows(.id = "area") |> 
  filter(to_id %in% bbox_h3_all) |>
  relocate(from_id, .before = everything()) |>
  cell_to_polygon(simple = FALSE) |> 
  st_as_sf() |> 
  tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Walk Travel Time to Town Centre")

#save tmap
tmap_save(tmap_ttm_walk, "Waikato Town Centre Access by Walking.html")
```


The Street network is an output included in all the maps.
```{r}
street_network <- r5r::street_network_to_sf(r5r_core_waikato)

street_network <- st_transform(street_network$edges, st_crs(waikato_bbox)) |>  
  filter(street_class %in% c("MOTORWAY", "SECONDARY", "TERTIARY")) #add "OTHER" if not mapping well

object.size(street_network) / 1000000
```


###########################################################
# calculate walking access to jobs and schools and save as an interactive map

- check the script will use the revised origin and destinations so that area cut off by water arent included
- 
```{r}
#load in previous data
#walk_results <- readRDS("walk_results.rds")

#cost to closest 1k jobs
threshold <- 1000 #considered lower threshold for walking but think this is still helpful, only Horotiu worked at 5k

calculate_access_to_jobs <- function(ttm, area){
  
  #add jobs to ttm
  ttm <- ttm %>%
    left_join(jobs_h3$jobs_data |> select(h3_address, jobs), by = join_by("to_id" == "h3_address"))
  
  ttm_threshold <- ttm %>%
  arrange(from_id, travel_time_p50) %>%
  group_by(from_id) %>%
  mutate(
    cum_jobs = cumsum(jobs)) %>%
  filter(cum_jobs >= threshold) %>%
  slice_min(order_by = travel_time_p50, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(from_id, min_time_to_1k_jobs = travel_time_p50, cum_jobs) |> #check from_id name is correct
  mutate(area = area)
  
  return(ttm_threshold)

}

ttm_walk_threshold_results <- purrr::map2(.x = ttm_walk_results_list,
            .y = names(ttm_walk_results_list), #what is the output from the ttms?
            .f = calculate_access_to_jobs)

#visually inspect ttm_walk_thresholds

ttm_walk_threshold_results|> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "YlGn", style = "cont")

#cost to closest school

calculate_access_to_school <- function(ttm, area, land_use_data, travel_cost = "travel_time_p50"){
  
  cost_to_closest <- cost_to_closest(ttm, 
                land_use_data = land_use_data,
                opportunity = "school", 
                travel_cost = travel_cost) |>
    mutate(travel_time_p50 = ifelse(travel_time_p50 == Inf, NA, travel_time_p50)) #inf results to NA
  
  return(cost_to_closest)
}

# is the opportunities data at the right resolution?

school_access_results <- purrr::map2(.x = ttm_walk_results_list, #it expects a list so no prep required
                                     .y = names(ttm_walk_results_list),
                                     .f = calculate_access_to_school,
                                     opportunities$combined_summary_all,
                                     .progress = TRUE)

#combine school results for mapping
tmap_school_access <- school_access_results |> bind_rows(.id = "area") |> relocate(id, .before = everything()) |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Walk Travel Time to Nearest School")

#save tmap
tmap_save(tmap_school_access, "Waikato School Access by Walking.html")

#combine jobs access results for mapping
tmap_jobs_access <- ttm_walk_threshold_results |> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "-YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time") + tm_layout(title = "Walk Travel Time to Nearest 1k Jobs")

#save tmap
tmap_save(tmap_jobs_access, "Waikato Jobs Access by Walking.html")

walk_results <- list(ttm_walk_threshold_results, school_access_results)

saveRDS(walk_results, "walk_results.rds")


```

#############################################################3
#create standardised walking maps to town centre and save

Finds the centroid of the bbox and filters the ttm to that zone. 
```{r}
static_population_mapping <- function(sf_data, colour = "travel_time_p50", legend_title, location, res, mode, r5r_core, bbox, variable = "population"){
  
  #find centroid h3 of destination
  centroid_h3 <- bbox |> 
    st_centroid() |> 
    point_to_cell(res = res, simple = TRUE)
  
  #combine travel times with job, income and population data
  ttm <- sf_data 
  
  ttm_to_centroid <- ttm |> 
    filter(to_id == centroid_h3) |> 
    left_join(pop_h3$pop_data |> 
                select(h3_address, population) |>
                st_drop_geometry(), 
              by = join_by("from_id" == "h3_address")) |> #interested in population at origin
    cell_to_polygon(simple = FALSE) %>%
    st_as_sf()
  
  axis_title <- paste(legend_title, "to", location, "by", mode, "at res", res, sep =  " ")
  
  street_network <- street_network  |> 
    st_crop(bbox)
  
  #convert to bbox format
  bbox <- bbox |> st_bbox()
  
  map <- ggplot(ttm_to_centroid) +
    geom_sf(aes(fill = !!sym(colour)), colour = NA, size = 0.01) +
    scale_fill_viridis_c(direction = 1) +
    labs(fill = legend_title) +
    ggtitle(axis_title) +
    theme_minimal() +
    theme(legend.position = "right",
          axis.text = element_blank(),  
          axis.ticks = element_blank()) +
    geom_sf(data = street_network, linewidth = 0.1, colour = "gray57") +
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
    )
  
  #create binned travel time to population data
  aggregated_data <- ttm_to_centroid %>%
    mutate(
      travel_time_bin = cut(
        travel_time_p50,
        breaks = seq(0, max(travel_time_p50, na.rm = TRUE), by = 5),
        right = FALSE,  # Left-closed, right-open intervals
        labels = paste(seq(0, max(travel_time_p50, na.rm = TRUE) - 5, by = 5), 
                       seq(5, max(travel_time_p50, na.rm = TRUE), by = 5),
                       sep = "-")
      )
    ) %>%
    group_by(travel_time_bin) %>%
    summarise(population = sum(population, na.rm = TRUE))
  
  # Plot
  legend_plot <- ggplot(aggregated_data,
                        aes(y = population, x = travel_time_bin, fill = travel_time_bin)) +
    geom_col(width = 1) +
    scale_fill_viridis_d(direction = 1) +
    labs(x = "Travel Time", y = "Population", fill = "Travel Time") +
    theme_minimal() +
    coord_flip()
  
  legend <- ggplotGrob(legend_plot)
 
# save map
  ggsave(paste0(axis_title, ".png"), plot = map, device = "png")
  
  #save the population histogram
  histogram_title <- paste(legend_title, "to", location, "by", mode, "histogram", sep =  " ")
  ggsave(paste0(histogram_title, ".png"), plot = legend, device = "png")
  
  #write.csv(ttm, paste(axis_title, ".csv", sep = ""))
  
  message("data saved to file")
  
  return(map)
}

#map over the function
maps_population_all <- purrr::pmap(
  list(
    sf_data = ttm_walk_results_list,
    location = names(ttm_walk_results_list),
    bbox = bbox_list
  ),
  function(sf_data, location, bbox) {
    static_population_mapping(
      sf_data = sf_data,
      colour = "travel_time_p50",
      legend_title = "Travel Time",
      location = location,
      res = res,
      mode = "WALK",
      r5r_core = r5r_core_waikato,
      bbox = bbox,
      variable = "Travel Time"
    )
  }
)


  #pretty_names <- your_list_of_strings %>%
  #gsub("_bbox$", "", .) %>% #removes bbox from the end. Global Substitution. data is final arg
  #gsub("_", " ", .) %>% #removes the underscores
  #stringr::str_to_title() #creates standard title names

#remove large data sets
rm(ttm_walk_results_list)
gc()
```

#Create standardised static maps for access to schools

A list of data frames concerning access to schools has already been created. We use a similar albeit simpler process to create these maps.
```{r}
static_school_mapping <- function(sf_data, colour = "travel_time_p50", legend_title, location, res, mode, r5r_core, bbox, variable = "population"){
  
    #combine travel times with job, income and population data
  ttm <- sf_data |> 
    filter(!is.na(travel_time_p50)) |>
    left_join(pop_h3$pop_data |> 
                select(h3_address, population) |>
                st_drop_geometry(), 
              by = join_by("id" == "h3_address")) |> #interested in population at origin
    cell_to_polygon(simple = FALSE) %>%
    st_as_sf()
  
  axis_title <- paste(legend_title, "to", location, "schools by", mode, "at res", res, sep =  " ")
  
  street_network <-  street_network |> 
    st_crop(bbox)
  
  #convert to bbox format
  bbox <- bbox |> st_bbox()
  
  map <- ggplot(ttm) +
    geom_sf(aes(fill = !!sym(colour)), colour = NA, size = 0.01) +
    scale_fill_viridis_c(direction = 1) +
    labs(fill = legend_title) +
    ggtitle(axis_title) +
    theme_minimal() +
    theme(legend.position = "right",
          axis.text = element_blank(),  
          axis.ticks = element_blank()) +
    geom_sf(data = street_network, linewidth = 0.1, colour = "gray57") +
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
    )
  
  #create binned travel time to population data
  aggregated_data <- ttm %>%
    mutate(
      travel_time_bin = cut(
        travel_time_p50,
        breaks = seq(0, max(travel_time_p50, na.rm = TRUE), by = 5),
        right = FALSE,  # Left-closed, right-open intervals
        labels = paste(seq(0, max(travel_time_p50, na.rm = TRUE) - 5, by = 5), 
                       seq(5, max(travel_time_p50, na.rm = TRUE), by = 5),
                       sep = "-")
      )
    ) %>%
    group_by(travel_time_bin) %>%
    summarise(population = sum(population, na.rm = TRUE))
  
  # Plot
  legend_plot <- ggplot(aggregated_data,
                        aes(y = population, x = travel_time_bin, fill = travel_time_bin)) +
    geom_col(width = 1) +
    scale_fill_viridis_d(direction = 1) +
    labs(x = "Travel Time", y = "Population", fill = "Travel Time") +
    theme_minimal() +
    coord_flip()
  
  legend <- ggplotGrob(legend_plot)
 
# save map
  ggsave(paste0(axis_title, ".png"), plot = map, device = "png")
  
  #save the population histogram
  histogram_title <- paste(legend_title, "to", location, "Schools", "by", mode, "histogram", sep =  " ")
  ggsave(paste0(histogram_title, ".png"), plot = legend, device = "png")
  
  #write.csv(ttm, paste(axis_title, ".csv", sep = ""))
  
  message("data saved to file")
  
  return(map)
}

```

# use for individual location
```{r}
maps_raglan <- static_school_mapping(
  sf_data      = school_access_results$raglan_bbox,
  colour       = "travel_time_p50",
  legend_title = "Travel Time",
  location     = "raglan",
  res          = res,
  mode         = "WALK",
  r5r_core     = r5r_core_waikato,
  bbox         = bbox_list$raglan_bbox,
  variable     = "Travel Time"
)

```


# Map over the function
```{r}

#map over the function.
maps_school_all <- purrr::pmap(
  list(
    sf_data = school_access_results,
    location = names(school_access_results),
    bbox = bbox_list
  ),
  function(sf_data, location, bbox) {
    static_school_mapping(
      sf_data = sf_data,
      colour = "travel_time_p50",
      legend_title = "Travel Time",
      location = location,
      res = res,
      mode = "WALK",
      r5r_core = r5r_core_waikato,
      bbox = bbox,
      variable = "Travel Time"
    )
  }
)
```


Now find access to bus stops

```{r}

# confirm location of bus stops
transit_network <- r5r::transit_network_to_sf(r5r_core_waikato)

transit_network <- transit_network$stops |>
  rowwise() |>
  filter(!st_is_empty(geometry)) |>
  st_transform(crs = 4326) |>
  point_to_cell(simple = FALSE, res = res) |>
  st_drop_geometry() |>
  mutate(bus_stop = 1) |> # identifier 
  select(bus_stop, h3_resolution_10)


#update land use data with bus stop locations using waikato routing cells

opportunities$combined_summary_all <- opportunities$combined_summary_all |>
  left_join(transit_network, by = join_by(id == h3_resolution_10)) |>
  mutate(across(
    where(is.numeric), #only apply to numeric columns
    ~ tidyr::replace_na(., 0)
  ))

raglan_bus_stop_access <- cost_to_closest(ttm_walk_results_list$raglan_bbox, 
                land_use_data = opportunities$combined_summary_all,
                opportunity = "bus_stop", 
                travel_cost = "travel_time_p50") |>
    mutate(travel_time_p50 = ifelse(travel_time_p50 == Inf, NA, travel_time_p50))
```
# use for individual location
```{r}

static_bus_stop_mapping <- function(sf_data, colour = "travel_time_p50", legend_title, location, res, mode, bbox, variable = "population"){
  
    #combine travel times with job, income and population data
  ttm <- sf_data |> 
    filter(!is.na(travel_time_p50)) |>
    left_join(pop_h3$pop_data |> 
                select(h3_address, population) |>
                st_drop_geometry(), 
              by = join_by("id" == "h3_address")) |> #interested in population at origin
    cell_to_polygon(simple = FALSE) %>%
    st_as_sf()
  
  axis_title <- paste(legend_title, "to", location, "bus stop by", mode, "at res", res, sep =  " ")
  
  street_network <-  street_network |> 
    st_crop(bbox)
  
  #convert to bbox format
  bbox <- bbox |> st_bbox()
  
  map <- ggplot(ttm) +
    geom_sf(aes(fill = !!sym(colour)), colour = NA, size = 0.01) +
    scale_fill_viridis_c(direction = 1) +
    labs(fill = legend_title) +
    ggtitle(axis_title) +
    theme_minimal() +
    theme(legend.position = "right",
          axis.text = element_blank(),  
          axis.ticks = element_blank()) +
    geom_sf(data = street_network, linewidth = 0.1, colour = "gray57") +
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]),
    )
  
  #create binned travel time to population data
  aggregated_data <- ttm %>%
    mutate(
      travel_time_bin = cut(
        travel_time_p50,
        breaks = seq(0, max(travel_time_p50, na.rm = TRUE), by = 5),
        right = FALSE,  # Left-closed, right-open intervals
        labels = paste(seq(0, max(travel_time_p50, na.rm = TRUE) - 5, by = 5), 
                       seq(5, max(travel_time_p50, na.rm = TRUE), by = 5),
                       sep = "-")
      )
    ) %>%
    group_by(travel_time_bin) %>%
    summarise(population = sum(population, na.rm = TRUE))
  
  # Plot
  legend_plot <- ggplot(aggregated_data,
                        aes(y = population, x = travel_time_bin, fill = travel_time_bin)) +
    geom_col(width = 1) +
    scale_fill_viridis_d(direction = 1) +
    labs(x = "Travel Time", y = "Population", fill = "Travel Time") +
    theme_minimal() +
    coord_flip()
  
  legend <- ggplotGrob(legend_plot)
 
# save map
  ggsave(paste0(axis_title, ".png"), plot = map, device = "png")
  
  #save the population histogram
  histogram_title <- paste(legend_title, "to", location, "bus stop by", mode, "histogram", sep =  " ")
  ggsave(paste0(histogram_title, ".png"), plot = legend, device = "png")
  
  #write.csv(ttm, paste(axis_title, ".csv", sep = ""))
  
  message("data saved to file")
  
  return(map)
}

bus_stop_raglan <- static_bus_stop_mapping(
  sf_data      = raglan_bus_stop_access |> filter(!is.na(travel_time_p50)),
  colour       = "travel_time_p50",
  legend_title = "Travel Time",
  location     = "raglan",
  res          = res,
  mode         = "WALK",
  bbox         = bbox_list$raglan_bbox,
  variable     = "Travel Time"
)

```

#detour analysis
First this is done for schools as the ttm already exists in a filtered format
```{r}

schools <- opportunities$combined_summary_all |> filter(school == 1)

raglan_school_detour <- ttm_walk_results_list$raglan_bbox |> 
  filter(to_id %in% schools$id) |>
  detour_ratio(1) |> # return full distribution
  arrange(desc(detour_ratio))  |>
  filter(travel_time_p50 > 5) # some short long detour journeys 

raglan_detour_origins <- raglan_school_detour |> 
  select(from_id, detour_ratio) |> 
  left_join(waikato_routing_cells, by = join_by(from_id == id)) |> 
  rename(id = from_id) |>
  filter(!is.na(lon)) # must be an issues with some of the point locations

raglan_detour_destinations <- raglan_school_detour |> 
  filter(from_id %in% raglan_detour_origins$id) |>
  select(to_id, detour_ratio) |> 
  left_join(waikato_routing_cells, by = join_by(to_id == id)) |> 
  rename(id = to_id) |>
  filter(!is.na(lon)) # must be an issues with some of the point locations
# Example: Load R5 network

r5r_core_waikato <- setup_r5(data_path = destfile, overwrite = FALSE)

raglan_school_detour_itineraries <- detailed_itineraries(
  r5r_core = r5r_core_waikato,
  origins =  raglan_detour_origins,
  destinations = raglan_detour_destinations,
  mode = "WALK",
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time
)

raglan_school_detour_itineraries <- raglan_school_detour_itineraries |>
  left_join(raglan_school_detour, by = join_by(from_id, to_id)) |>
  filter(detour_ratio > 2)

tmap_mode("view")

tmap_raglan_school_detour_itineraries <- tm_shape(raglan_school_detour_itineraries[3,]) + tm_lines(col = "detour_ratio", lwd = 3)
```


################################################################
#cycling time matrix and access to jobs and education
#################################################################
#cycling based
```{r}
#load previously calculated ttm
#ttm_bicycle_results_list <- readRDS("ttm_bicycle_results_list.rds")

run_ttm_for_area <- function(study_area, wider_study_area, area_id) {
  
  #crop destinations 
  cropped_destinations <- waikato_routing_cells[
  st_intersects(waikato_routing_cells, study_area, sparse = FALSE),]
  
  #crop origins
  cropped_origins <- waikato_routing_cells[
  st_intersects(waikato_routing_cells, wider_study_area, sparse = FALSE),]
  
  #catch an issues with origin data
  if (nrow(cropped_destinations) == 0) {
    warning(paste("No origins found for area", area_id))
    return(NULL)
  }
  
  #run travel time matrix
  ttm <- travel_time_matrix(
    r5r_core_waikato,
    origins = cropped_origins,
    destinations = cropped_destinations,
    mode = "BICYCLE", #change mode independently of global settings
    departure_datetime = departure_datetime,
    max_walk_time = 10,
    max_trip_duration = 45,
    walk_speed = walk_speed
  )
 
  return(ttm)
}

#loop of list of bboxs
ttm_bicycle_results_list <- purrr::pmap(list(st_geometry(combined_bboxs),
                                          st_geometry(combined_bboxs_30km), 
                                          combined_bboxs$area_name),
  .f = run_ttm_for_area) |> set_names(combined_bboxs$area_name)

#save the ttm
saveRDS(ttm_bicycle_results_list, "ttm_bicycle_results_list.rds")

```

#map and save interactive cycling ttm map
```{r}

tmap_ttm_bicycle <- ttm_bicycle_results_list |> 
  bind_rows(.id = "area") |> 
  filter(to_id %in% bbox_h3_all) |>
  relocate(from_id, .before = everything()) |>
  cell_to_polygon(simple = FALSE) |> 
  st_as_sf() |> 
  tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Cycle Travel Time to Town Centre")

#save tmap
tmap_save(tmap_ttm_bicycle, "Waikato Town Centre Access by Cycling.html")


```

#cycling access
```{r}
#cost to closest 1k jobs
threshold <- 1000 

#calculate jobs function is already defined

#map over inputs
ttm_bicycle_threshold_results <- purrr::map2(.x = ttm_bicycle_results_list,
            .y = names(ttm_bicycle_results_list), #what is the output from the ttms?
            .f = calculate_access_to_jobs)

#visually inspect ttm_cycle_thresholds

ttm_bicycle_threshold_results|> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "YlGn", style = "cont")

#cost to closest school function already defined

school_bicycle_access_results <- purrr::map2(.x = ttm_bicycle_results_list, #it expects a list so no prep required
                                     .y = names(ttm_bicycle_results_list),
                                     .f = calculate_access_to_school,
                                     opportunities$combined_summary,
                                     .progress = TRUE)

```

#save cycling access results
```{r}

bicycling_results <- list(ttm_bicycle_threshold_results, school_bicycle_access_results)

saveRDS(bicycling_results, "bicycle_results.rds")

#combine school results for mapping
tmap_bicycle_school_access <- school_bicycle_access_results |> bind_rows(.id = "area") |> relocate(id, .before = everything()) |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Cycle Travel Time to Nearest School")

#save tmap
tmap_save(tmap_bicycle_school_access, "Waikato School Access by Cycling.html")

#combine jobs access results for mapping
tmap_bicycle_jobs_access <- ttm_bicycle_threshold_results |> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "-YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time") + tm_layout(title = "Bicycle Travel Time to Nearest 1k Jobs")

#save tmap
tmap_save(tmap_bicycle_jobs_access, "Waikato Jobs Access by Cycling.html")

#remove large data sets

```

#create standardised cycling maps and save
```{r}
#static population mapping function already defined under walking

#map over the function
maps_population_cycling_all <- purrr::pmap(
  list(
    sf_data = ttm_bicycle_results_list,
    location = names(ttm_bicycle_results_list),
    bbox = bbox_list
  ),
  function(sf_data, location, bbox) {
    static_population_mapping(
      sf_data = sf_data,
      colour = "travel_time_p50",
      legend_title = "Travel Time",
      location = location,
      res = res,
      mode = "BICYCLE",
      r5r_core = r5r_core_waikato,
      bbox = bbox,
      variable = "Travel Time"
    )
  }
)
rm(ttm_bicycle_results_list)
gc()
```

###########################
#transit based analysis
############################

#some pre-checks
```{r}
r5r::check_network_status(r5r_core_waikato)

#looks like a gtfs issue

#run gtfs validator?
```


Transit analysis is still failing due to java indexoutofboundsexception.

Based on what I saw earlier, could it be a lack of spatial information with some stops in the gtfs?

```{r}
#custom function to run ttm for all study areas defined under walking

#loop of list of bboxs
ttm_transit_results_list <- purrr::pmap(list(st_geometry(combined_bboxs),
                                          st_geometry(combined_bboxs_30km), 
                                          combined_bboxs$area_name),
                                        rep("TRANSIT", 11),
  .f = run_ttm_for_area) |> set_names(combined_bboxs$area_name)

saveRDS(ttm_transit_results_list, "ttm_transit_results_list.rds")

#run for single location
raglan_transit_ttm <- run_ttm_for_area(st_geometry(combined_bboxs[1,]),
                                       st_geometry(combined_bboxs_30km[1,]),
                                       combined_bboxs$area_name[1],
                                       "TRANSIT")
```

#map and save interactive transit ttm map
```{r}
tmap_ttm_transit <- ttm_transit_results_list |> 
  bind_rows(.id = "area") |> 
  filter(to_id %in% bbox_h3_all) |>
  relocate(from_id, .before = everything()) |>
  cell_to_polygon(simple = FALSE) |> 
  st_as_sf() |> 
  tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Public Transport Travel Time to Town Centre")

#save tmap
tmap_save(tmap_ttm_transit, "Waikato Town Centre Access by Public Transport.html")
```


#Transit access
```{r}
#cost to closest 1k jobs
threshold <- 1000 

#calculate jobs function is already defined

#map over inputs
ttm_transit_threshold_results <- purrr::map2(.x = ttm_transit_results_list,
            .y = names(ttm_transit_results_list), #what is the output from the ttms?
            .f = calculate_access_to_jobs)

#cost to closest school function already defined

school_transit_access_results <- purrr::map2(.x = ttm_transit_results_list, #it expects a list so no prep required
                                     .y = names(ttm_transit_results_list),
                                     .f = calculate_access_to_school,
                                     opportunities$combined_summary,
                                     .progress = TRUE)
```
#save transit access results
```{r}
transit_results <- list(ttm_transit_threshold_results, school_transit_access_results)

saveRDS(transit_results, "transit_results.rds")

#combine school results for mapping
tmap_transit_school_access <- school_transit_access_results |> bind_rows(.id = "area") |> relocate(id, .before = everything()) |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Public Transport Travel Time to Nearest School")

#save tmap
tmap_save(tmap_transit_school_access, "Waikato School Access by Public Transport.html")

#combine jobs access results for mapping
tmap_transit_jobs_access <- ttm_transit_threshold_results |> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "-YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time") + tm_layout(title = "Public Transport Travel Time to Nearest 1k Jobs")

#save tmap
tmap_save(tmap_transit_jobs_access, "Waikato Jobs Access by Public Transport.html")

#remove large data sets

```

#create standardised transit maps and save
```{r}
#static population mapping function already defined under walking

#map over the function
maps_population_transit_all <- purrr::pmap(
  list(
    sf_data = ttm_transit_results_list,
    location = names(ttm_transit_results_list),
    bbox = bbox_list
  ),
  function(sf_data, location, bbox) {
    static_population_mapping(
      sf_data = sf_data,
      colour = "travel_time_p50",
      legend_title = "Travel Time",
      location = location,
      res = res,
      mode = "TRANSIT",
      r5r_core = r5r_core_waikato,
      bbox = bbox,
      variable = "Travel Time"
    )
  }
)
rm(ttm_transit_results_list)
gc()
```

################################################################
#Car travel time matrix and access to jobs and education
#################################################################
#error catching verions
```{r}
#rm(ttm_car_results_list)
#ttm_car_results_list <- readRDS("ttm_car_results_list.rds")

ttm_sparse_list <- ttm_car_results_list %>%
  map(~ .x %>%
         tidyr::pivot_wider(
           names_from = to_id,
           values_from = travel_time_p50,
           values_fill = 0  # or use Inf if that’s your default non-connection
         ) %>%
         {
           df <- .
           rownames(df) <- df$from_id
           mat <- as.matrix(select(df, -from_id))
           Matrix::Matrix(mat, sparse = TRUE)
         }
     )




```

#Car based
```{r}
#ttm_car_results_list.rds <- readRDS("ttm_car_results_list.rds")

#load previously calculated ttm
#ttm_bicycle_results_list <- readRDS("ttm_bicycle_results_list.rds")

run_ttm_for_area <- function(study_area, wider_study_area, area_id) {
  
  cat("\nStarting area:", area_id, "\n")
  
  # crop destinations
  cropped_destinations <- waikato_routing_cells[
    st_intersects(waikato_routing_cells, study_area, sparse = FALSE),]
  
  cat("Cropped destinations for", area_id, ": ", nrow(cropped_destinations), "cells\n")
  
  # crop origins
  cropped_origins <- waikato_routing_cells[
    st_intersects(waikato_routing_cells, wider_study_area, sparse = FALSE),]
  
  cat("Cropped origins for", area_id, ": ", nrow(cropped_origins), "cells\n")
  
  # catch any issues with destination or origin data
  if (nrow(cropped_destinations) == 0) {
    warning(paste("No destinations found for area", area_id))
    return(NULL)
  }
  
  if (nrow(cropped_origins) == 0) {
    warning(paste("No origins found for area", area_id))
    return(NULL)
  }
  
  cat("Running travel_time_matrix for", area_id, "...\n")
  
  # run travel time matrix with error catching
  ttm <- tryCatch(
    {
      travel_time_matrix(
        r5r_core_waikato,
        origins = cropped_origins,
        destinations = cropped_destinations,
        mode = "CAR",  # or "TRANSIT", "WALK", etc
        departure_datetime = departure_datetime,
        max_walk_time = 5,
        max_trip_duration = 30,
        walk_speed = walk_speed
      )
    },
    error = function(e) {
      message(paste("Error in area", area_id, ":", e$message))
      return(NULL)
    }
  )
  
  if (!is.null(ttm)) {
    cat("Finished travel_time_matrix for", area_id, "\n")
  } else {
    cat("Failed travel_time_matrix for", area_id, "\n")
  }
  
  return(ttm)
}

#loop of list of bboxs
ttm_car_results_list <- purrr::pmap(list(st_geometry(combined_bboxs),
                                          st_geometry(combined_bboxs_30km), 
                                          combined_bboxs$area_name),
  .f = run_ttm_for_area) |> set_names(combined_bboxs$area_name)

#save the ttm
saveRDS(ttm_car_results_list, "ttm_car_results_list.rds", compress = "xz")

```

#map and save interactive car ttm map
```{r}
#once ttm re run with routing cells then drop the additional raglan point
tmap_mode("view")



#ttm_car_results_list$raglan_bbox$to_id |> unique() |> cell_to_point() |> tm_shape() + tm_dots()
#waikato_routing_cells |> st_crop(bbox_list$raglan_bbox) |> tm_shape() + tm_dots()

#custom function filter and create travel time multipolygons rather than many individual polygons
filter_large_ttm_list <- function(ttm){
  ttm <- ttm |>
    filter(to_id %in% bbox_h3_all) |> #filters to only town centres as destinations
    relocate(from_id, .before = everything()) |>
    mutate(tt_bin = cut(travel_time_p50, breaks = seq(0, 60, by = 5))) |>
    group_by(tt_bin) |>
    summarise(geometry = cells_to_multipolygon(from_id)) |>
    st_as_sf()
  
    return(ttm)
}
#map over function
ttm_car_town_centres <- purrr::map(.x = ttm_car_results_list, .f = filter_large_ttm_list) |>
  bind_rows(.id = "area") #collapse list into single object

#test map
tmap_mode("plot")
tm_shape(ttm_car_town_centres) + tm_polgons(col = "tt_bin", palette = "viridis")

#test on single df

test_car <- ttm_car_results_list$n

|> #filters to only town centres as destinations
  relocate(from_id, .before = everything()) |>
  mutate(tt_bin = cut(travel_time_p50, breaks = seq(0, 60, by = 5))) |>
  group_by(tt_bin) |>
  summarise(geometry = cells_to_multipolygon(from_id)) |>
  st_as_sf()

tm_shape(test_car) + tm_polygons(col = "tt_bin", palette = "viridis")

ttm_car_results_list.rds$raglan_bbox %>% 
  mutate(id = trimws(from_id)) %>% 
  filter(is.na(id) | nchar(id) != 15 | grepl("[^0-9a-f]", id)) %>% 
  distinct(id) %>% 
  pull(id) %>% 
  head(20)



tmap_ttm_car <- ttm_car_results_list |> 
  bind_rows(.id = "area") |> 
  filter(to_id %in% bbox_h3_all) |> #filters to only town centres as destinations
  relocate(from_id, .before = everything()) |>
  cell_to_polygon(simple = FALSE) |> 
  st_as_sf() |> 
  mutate(tt_bin = cut(travel_time_p50, breaks = seq(0, 60, by = 5))) %>%
  group_by(tt_bin) %>%
  summarise(geometry = st_union(geometry)) %>% #combines polygons in the same bin
  ungroup() %>%
  tm_shape() + tm_polygons(col = "tt_bin", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Car Travel Time to Town Centre")

tmap_ttm_car <- ttm_car_results_list |> 
  bind_rows(.id = "area") |> #filter to_id first
  filter(to_id %in% bbox_h3_all) |> #filters to only town centres as destinations
  relocate(from_id, .before = everything()) |>
  mutate(tt_bin = cut(travel_time_p50, breaks = seq(0, 60, by = 5)))

#ms_simplify(keep = 0.05) |> #removes unnecersary geometry to retain 5%
#save tmap
tmap_save(tmap_ttm_car, "Waikato Town Centre Access by Car.html")

```

#car access
```{r}
#cost to closest 1k jobs
threshold <- 1000 

#calculate jobs function is already defined

#map over inputs
ttm_car_threshold_results <- purrr::map2(.x = ttm_car_results_list,
            .y = names(ttm_car_results_list), #what is the output from the ttms?
            .f = calculate_access_to_jobs)

#visually inspect ttm_cycle_thresholds

ttm_car_threshold_results|> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "YlGn", style = "cont")

#cost to closest school function already defined

school_car_access_results <- purrr::map2(.x = ttm_car_results_list, #it expects a list so no prep required
                                     .y = names(ttm_car_results_list),
                                     .f = calculate_access_to_school,
                                     opportunities$combined_summary,
                                     .progress = TRUE)

```

#save car access results
```{r}

car_results <- list(ttm_car_threshold_results, school_car_access_results)

saveRDS(car_results, "car_results.rds")

#combine school results for mapping
tmap_car_school_access <- school_car_access_results |> bind_rows(.id = "area") |> relocate(id, .before = everything()) |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "travel_time_p50", alpha = 0.7, palette = "YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time P50") + tm_layout(title = "Car Travel Time to Nearest School")

#save tmap
tmap_save(tmap_car_school_access, "Waikato School Access by Car.html")

#combine jobs access results for mapping
tmap_car_jobs_access <- ttm_car_threshold_results |> bind_rows(.id = "area") |> cell_to_polygon(simple = FALSE) |> st_as_sf() |> tm_shape() + tm_polygons(col = "min_time_to_1k_jobs", alpha = 0.7, palette = "-YlGn", style = "cont", textNA = ">45 minutes", colorNA = "#74b3eb", lwd = 0, title = "Travel Time") + tm_layout(title = "Car Travel Time to Nearest 1k Jobs")

#save tmap
tmap_save(tmap_car_jobs_access, "Waikato Jobs Access by Car.html")


```

#create standardised car maps and save


##################################
# Big loop for car access w progress checker
##################################

```{r}
process_area <- function(
  area_id,
  study_area,       # tight bbox for destinations
  wider_area,       # wider bbox for origins
  mode,             # "WALK", "BICYCLE", "CAR", or "TRANSIT"
  ttm_dir,          # where to save raw TTM .rds
  out_dir           # where to save maps & access .rds
){
  cat("\n==== Processing area:", area_id, "====\n")
  
  # 1. crop origins & dests
  origins <- waikato_routing_cells[
    st_intersects(waikato_routing_cells, wider_area, sparse = FALSE), ]
  
  dests   <- waikato_routing_cells[
    st_intersects(waikato_routing_cells, study_area, sparse = FALSE), ]
  
  cat("  • Origins:", nrow(origins), "cells\n")
  cat("  • Destinations:", nrow(dests), "cells\n")
  
  if(nrow(origins)==0 || nrow(dests)==0){
    warning("No origins/dests for ", area_id)
    return(NULL)
  }
  
 
  # 2. run travel_time_matrix with error catching
  cat("  • Running travel_time_matrix(", mode, ") …\n")
  ttm <- tryCatch(
    {
      travel_time_matrix(
        r5r_core_waikato,
        origins            = origins,
        destinations       = dests,
        mode               = mode,
        departure_datetime = departure_datetime,
        time_window = 1, #for car it doesnt matter
        max_walk_time      = 3, #limit off grid walking
        max_trip_duration  = 30,
        walk_speed         = 3.6 #reduce walk speed to more realistic speed.
      )
    },
    error = function(e) {
      message("    ✗ Error in travel_time_matrix for ", area_id, ": ", e$message, "\n")
      return(NULL)
    }
  )
  if(is.null(ttm)){
    cat("    → ttm FAILED for", area_id, "\n")
    return(NULL)
  } else {
    cat("    ✓ ttm complete for", area_id, "\n")
  }
  
  # save raw TTM
  ttm_file <- file.path(ttm_dir, paste0("ttm_", tolower(mode), "_", area_id, ".rds"))
  saveRDS(ttm, ttm_file)
  cat("  • Saved raw TTM to", ttm_file, "\n")
  
  # 3. build multipolygon map
  cat("  • Building multipolygon bins …\n")
  
  bins <- ttm %>%
    filter(to_id %in% bbox_h3_all) %>%
    mutate(tt_bin = cut(travel_time_p50, seq(0, 60, by = 1))) %>%
    group_by(tt_bin) %>%
    summarise(h3 = list(unique(from_id)), .groups="drop") %>%
    rowwise() %>%
    mutate(geometry = cells_to_multipolygon(h3, simple=TRUE)) %>%
    st_as_sf(crs=st_crs(waikato_routing_cells))
  
  mapfile  <- file.path(out_dir, paste0("map_ttm_", tolower(mode), "_", area_id, ".html"))
  tmap_mode("view")
  map_obj <- tm_shape(bins) + tm_polygons(col = "tt_bin", palette="-viridis", alpha=0.5, lwd = 0, title = "Travel Time", textNA = "> 30 Minutes")
  
  tmap_save(map_obj, file.path(out_dir, paste0("map_ttm_", tolower(mode), "_", area_id, ".html")))
  
  cat("  • Saved interactive map to", mapfile, "\n")
  
  # 4. compute 1k-jobs access
  cat("  • Calculating 1k-jobs access …\n")
  
  jobs_access <- ttm %>%
    left_join(
      jobs_h3$jobs_data %>% select(h3_address, jobs),
      by = c(to_id = "h3_address")
    ) %>%
    arrange(from_id, travel_time_p50) %>%
    group_by(from_id) %>%
    mutate(cum_jobs = cumsum(jobs)) %>%
    filter(cum_jobs >= 1000) %>%
    slice_min(travel_time_p50, with_ties=FALSE) %>%
    ungroup() %>%
    transmute(
      from_id,
      min_time_to_1k_jobs = travel_time_p50,
      cum_jobs
    )
  
  jobs_file <- file.path(out_dir, paste0("jobs_access_", area_id, ".rds"))
  saveRDS(jobs_access, jobs_file)
  cat("  • Saved jobs access to", jobs_file, "\n")
  
  # 5. compute nearest-school access
  cat("  • Calculating school access …\n")
  
  school_access <- calculate_access_to_school(ttm, area_id, land_use_data = opportunities$combined_summary_all)
  
  school_file <- file.path(out_dir, paste0("school_access_", area_id, ".rds"))
  
  saveRDS(school_access, school_file)
  
  cat("  • Saved school access to", school_file, "\n")
  
  # 6. summary return
  cat("==== Done with area:", area_id, "====\n")
  
  #return key details and access data, but not the ttm
  return(list(
    area        = area_id,
    ttm_file    = ttm_file,
    map_file    = mapfile,
    jobs_file   = jobs_file,
    school_file = school_file,
    school_access = school_access,
    jobs_access = jobs_access)
  )
}

# Example: run across all areas with progress bar
library(progressr)
handlers("txtprogressbar")  # or "rstudio" if you're in RStudio

results <- with_progress({
  p <- progressor(steps = nrow(combined_bboxs))
  purrr::pmap(
    list(
      area_id    = combined_bboxs$area_name,
      study_area = st_geometry(combined_bboxs),
      wider_area = st_geometry(combined_bboxs_30km)
    ),
    function(area_id, study_area, wider_area){
      p(sprintf("Area: %s", area_id))
      process_area(
        area_id    = area_id,
        study_area = study_area,
        wider_area = wider_area,
        mode       = "CAR",
        ttm_dir    = "C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato",
        out_dir    = "C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato"
      )
    }
  )
})

#car_test <- readRDS("C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato/ttm_car_raglan_bbox.rds")

```

```{r}
#static population mapping function already defined under walking

#read in the ttm data
results |> bind_rows() |> View()
results |> bind_rows() |> select(map_file)

ttm_car_results_list <- list(raglan = readRDS("ttm_car_raglan_bbox.rds"))

#map over the function
maps_population_car_all <- purrr::pmap(
  list(
    sf_data = ttm_car_results_list,
    location = names(ttm_car_results_list),
    bbox = bbox_list
  ),
  function(sf_data, location, bbox) {
    static_population_mapping(
      sf_data = sf_data,
      colour = "travel_time_p50",
      legend_title = "Travel Time",
      location = location,
      res = res,
      mode = "CAR",
      r5r_core = r5r_core_waikato,
      bbox = bbox,
      variable = "Travel Time"
    )
  }
)
#remove large data sets
rm(ttm_car_results_list)
gc()
```



############################################
#Travel time: access to jobs and education
############################################



#cycling access
```{r}

```

#transit access
```{r}

```

#car access
```{r}

```

############################
#agglomeration improvements
############################



#map
```{r}
tmap <- tmap_access_plot(ttm_waikato, "Waikato TTM by PT", "travel_time_p50", "travel_time_p50")
```


Allocate to hex grid
```{r}
waikato_health_centroid <- waikato_health$osm_polygons %>% 
  st_centroid() %>%
  point_to_cell(simple = FALSE, res = res)  # 

# %>%
  rename(id = h3_resolution_7) %>%
  mutate(opportunities = 1)

waikato_health_centroid <- left_join(waikato_h3_id, waikato_health_centroid[1:3,])

library(tidyr)
waikato_health_centroid <- mutate(waikato_health_centroid, opportunities = replace_na(opportunities, 0))

```

cumulative hospitals
```{r}
# rename column to use {accessibility} package
#data.table::setnames(ttm_waikato, "travel_time_p50", "travel_time")



cum_opportunities <- cumulative_cutoff(
  ttm_waikato,
  land_use_data = waikato_health_centroid,
  opportunity = "opportunities",
  travel_cost = "travel_time",
  cutoff = 45)


```

map this
```{r}
ttm_waikato_opps <- cell_to_polygon(cum_opportunities, simple = FALSE)

waikato_opps_poly <- st_as_sf(ttm_waikato_opps)

tmap_waikato_opps <- tm_shape(waikato_opps_poly) + tm_polygons(col = "opportunities", style = "pretty" , alpha = 0.5, palette = "viridis") + tm_borders(alpha = 0.0)

tmap_mode("view")

tmap_waikato_opps #+ tmap_h2s_stops


```
A multicriteria asessement of 


# download road network
##consider updating and using dodgr for routing purposes
```{r}

#waikato_roads <- opq(waikato_matrix) %>% # Waikato, NZ
 #   add_osm_feature (key = "highway", value_exact = FALSE) %>%
    osmdata_sf ()
waikato_roads
```
cost to closest
```{r}
opportunity <- "school"

land_use_data <- filter(opportunities$combined_summary, !!sym(opportunity) > 0)
  
waikato_school_closest <- cost_to_closest(
  ttm_waikato,
  land_use_data = land_use_data,
  opportunity = opportunity,
  travel_cost = "travel_time_p50"
)

waikato_school_closest <- waikato_school_closest %>% left_join(waikato_pop$localised_population_interp, by = c("id" = "h3")) %>%
  select(-geom) %>%
  tidyr::replace_na(waikato_school_closest, list("population" = 0))

waikato_school_closest_summary <- waikato_school_closest %>%
  filter(population > 0)

ggplot(waikato_school_closest_summary, aes(x = travel_time_p50, y = population)) + 
  geom_point()

summary(waikato_school_closest_summary)
```


```{r}
tmap_access_plot_id(waikato_school_closest, "Waikato School Access by PT", "Waikato School Access by PT.html", "travel_time_p50", "travel_time_p50")
```



map cost to closest
```{r}
ttm_waikato_hospital_closest <- cell_to_polygon(waikato_hospital_closest, simple = FALSE)

waikato_hospital_closest_poly <- st_as_sf(ttm_waikato_hospital_closest)

waikato_hospital_closest_poly <- st_centroid(waikato_hospital_closest_poly)

tmap_waikato_hospital_closest_poly <- tm_shape(waikato_hospital_closest_poly) + tm_polygons(col = "travel_time", style = "pretty" , alpha = 0.5, palette = "viridis", lwd = 0.01, title = "Travel Time (gravity)") + tm_layout(title = "Hospital Access in Waikato by Car")

tmap_mode("view")

tmap_waikato_hospital_closest_poly #+ tmap_h2s_stops

tmap_save(tmap_waikato_hospital_closest_poly, "Travel time to closest hospital.html")
```


Gravity decay function to closest
```{r}
waikato_hospital_closest_gravity <- gravity(
  ttm_waikato,
  land_use_data = waikato_health_centroid,
  opportunity = "opportunities",
  travel_cost = "travel_time",
  decay_function = decay_exponential(0.01)
)
```

Map hospital decay function

```{r}
waikato_hospital_closest_gravity_poly <- cell_to_polygon(waikato_hospital_closest_gravity, simple = FALSE)

waikato_hospital_closest_gravity_poly <- st_as_sf(waikato_hospital_closest_gravity_poly)

tmap_waikato_hospital_closest_gravity_poly <- tm_shape(waikato_hospital_closest_gravity_poly) + tm_polygons(col = "opportunities", style = "pretty" , alpha = 0.5, palette = "viridis", lwd = 0.01, title = "Travel Time (gravity)") + tm_layout(title = "Hospital Access in Waikato by Car (gravity decay")

tmap_mode("view")

tmap_waikato_hospital_closest_gravity_poly #+ tmap_h2s_stops

tmap_save(tmap_waikato_hospital_closest_poly, "Travel time to closest hospital.html")
```

Allocate to hex grid
```{r}
waikato_shops_centroid <- waikato_shops$osm_polygons %>% 
  st_centroid() %>%
  point_to_cell(simple = FALSE, res = res) %>%
  rename(id = h3_resolution_7) %>%
  mutate(opportunities = 1)

waikato_shops_centroid <- left_join(waikato_h3_id, waikato_shops_centroid)

library(tidyr)
waikato_shops_centroid <- mutate(waikato_shops_centroid, opportunities = replace_na(opportunities, 0))

```

cost to closest groceries

```{r}
waikato_shops_closest <- cost_to_closest(
  ttm_waikato,
  land_use_data = waikato_shops_centroid,
  opportunity = "opportunities",
  travel_cost = "travel_time"
)
```


map the groceries
```{r}
ttm_waikato_shops_closest <- cell_to_polygon(waikato_shops_closest, simple = FALSE)

waikato_shops_closest_poly <- st_as_sf(ttm_waikato_shops_closest)

tmap_waikato_shops_closest <- tm_shape(waikato_shops_closest_poly) + tm_polygons(col = "travel_time", style = "pretty" , alpha = 0.5, palette = "viridis", lwd = 0.01)

tmap_mode("view")

tmap_waikato_shops_closest #+ tmap_h2s_stops

tmap_save(tmap_waikato_shops_closest, "Travel time to closest grocery store by car.html")
```
Gravity decay for shops
```{r}
waikato_shops_closest_gravity <- gravity(
  ttm_waikato,
  land_use_data = waikato_shops_centroid,
  opportunity = "opportunities",
  travel_cost = "travel_time",
  decay_function = decay_exponential(0.08)
)
```
Map groceries decay function

```{r}
waikato_shops_closest_gravity_poly <- cell_to_polygon(waikato_shops_closest_gravity, simple = FALSE)

waikato_shops_closest_gravity_poly <- st_as_sf(waikato_shops_closest_gravity_poly)

tmap_waikato_shops_closest_gravity_poly <- tm_shape(waikato_shops_closest_gravity_poly) + tm_polygons(col = "opportunities", style = "pretty" , alpha = 0.5, palette = "viridis", lwd = 0.01, title = "Travel Time (gravity decay)") + tm_layout(title = "Grocery Store Access in Waikato by Car (using a gravity decay of 0.08")

tmap_mode("view")

tmap_waikato_shops_closest_gravity_poly #+ tmap_h2s_stops

tmap_save(tmap_waikato_shops_closest_gravity_poly, "Grocery Store Access in Waikato (gravity decay function).html")
```
################3
#archive
##############
#Travel time matrix for all to some. If we do all to all that is over 7 billion pairs
```{r}

#filter the destinations for just the ones we want
#destinations <- opportunities$combined_summary["school" = c(1,2),]

#check if already written to disk and date

#fix any errors with inputs
waikato_routing_cells <- waikato_routing_cells %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  filter(!st_is_empty(.)) %>%
  mutate(id = as.character(id))


#currently set up so that its all to all 
ttm_waikato_pt <- travel_time_matrix(
  r5r_core_waikato,
  origins = waikato_routing_cells |> st_crop(bbox_list[[1]]) |> select(-id, -lat, -lon) |> class(),
  destinations = waikato_routing_cells |> select(-id, -lat, -lon),
  mode = mode,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  max_trip_duration = max_trip_duration,
  walk_speed = walk_speed
)

head(ttm_waikato_pt)

saveRDS(ttm_waikato_pt, "ttm_waikato_pt.rds")

#save results to disk
```
#check gtfs data
```{r}
gtfs <- gtfstools::read_gtfs("C:/Users/ellio/OneDrive/Documents/GIS Data/R5R/Waikato/Router_do_min/improved-gtfs-waikato-regional-council-1.zip")



gtfstools::validate_gtfs(gtfs, tempfile("validation_result"), gtfstools::download_validator(tempdir()))
```

#rnatural earth boundaries
```{r}
#provides poor boundaries that overlaps with land

#download the Natural Earth NZ polygon (everything that’s land)
land_polys <- rnaturalearth::ne_countries(scale = 10, country = "New Zealand") |> st_crop(waikato_bbox)
            # optional: limit to your bbox

#find h3 cells on land
land_cells <- polygon_to_cells(land_polys, res = res, simple = TRUE) |> 
  list_c() #flatten list

sub_waikato_routing_cells <- all_waikato_routing_cells[all_waikato_routing_cells$id %in% land_cells,]

```


